Ultrafast Electron Microscopes: Design Criteria, Electron Sources, and Column Modeling

### 2013-04-11 inc/photocathode/este/este.tex
citations are now preceeded by non-breaking space (~)

Recent experiments in our group have come to change our understanding of several photoemission processes, once thought to be well understood \cite{berger_excited_2012}.
Indeed, while this knowledge has been commonplace in other fields, such as angle-resolved photoemission spectroscopy (ARPES) \cite{himpsel_angle-resolved_1983}, to our knowledge the effect has not been considered for electron beam generation.
In this section, we show that, at least for thermionic emission, the effective mass $m^*$ of the electronic state from which the electron was emitted affects the rms transverse momentum; that is, the expression for $\Delta p_{\smallT}$ developed in Refs. \cite{dowell_quantum_2009} and \cite{jensen_emittance_2010} for thermionic electron emission should be rewritten to read $\sqrt{m^* k_B T_e}$, where $T_e$ is the temperature of the emitting electron distribution and $k_B$ is Boltzmann's constant.
Direct observation of the transverse momentum distribution for excited-state thermionic emission (ESTE) from GaSb and InSb, two similar III-V zincblende semiconductors, demonstrates the $m^*$ dependence.
Simulations of the experiment using our extended analytical Gaussian (AG) model of electron pulse propagation (see Section \ref{sec:external_forces}) are in close agreement with the data.

@@

\ref{fig:este-semicond} depicts the results obtained for GaSb and InSb photocathode materials under pulsed UV (261nm) laser irradiation using the same experimental technique employed for PAPE, 2PTE, and single-photon PE studies.
Both samples are cut from [100]-oriented polished wafers; the GaSb is undoped and the InSb is moderately $p$-type.
Prior studies \cite{gobeli_photoelectric_1965} indicate that single-photon photoemission should not possible for either semiconductor --- the effective photoemission work functions (about 4.8eV for undoped GaSb and 4.89eV for $p$-type InSb (Fermi level pinned at valence band maximum)) being greater than the 4.75eV photon energy even when the 34meV Schottky barrier suppression due to the applied 8kV/cm DC field is included \cite{dowell_quantum_2009}.
Nonetheless, for both zincblende semiconductors, significant laser-driven emission (more than expected for a single-photon field emission process) is observed with a yield nearly linearly proportional to the $\sim$4ps UV laser pulse energy.
Note that due to the surprisingly large yield, we repeated the measurements, this time replacing the the scintillator screen with a Faraday cup and repeating the laser energy sequence.
In this way, the yields for these photocathodes were measured directly and absolutely, unlike the results presented in the preceeding sections, for which relative measurements were sufficient.

The known optical properties of the two semiconductors \cite{aspnes_dielectric_1983} indicate a strong absorption at 261nm, with an optical absorption depth of only 7-8nm, which, based on the band structures of GaSb \cite{chelikowsky_nonlocal_1976} and InSb \cite{chelikowsky_erratum_1984}, is primarily due to promotion of electrons near the $\Gamma$ point from the valence band (heavy-hole, light-hole, and split-off bands) directly into the upper $\Gamma_8$ conduction band located at 3.77eV and 3.59eV above the valence band maximum in GaSb and InSb, respectively.
Assuming parabolic bands and an estimated effective mass $m^*$ of about 0.3$m$ (0.5$m$) for the $\Gamma_8$ conduction band in GaSb (InSb), we determine that these electrons are excited with an average excess energy (above the $\Gamma_8$ conduction band minimum) of $\sim$0.35eV ($\sim$0.41eV); that is, an initial electron temperature $T_e$ of $\sim$4,200K ($\sim$4,900K).
For the $\sim$0.1nJ incident UV pulse energies, $\sim40\mu$m laser spot size, and $p$-polarized reflectance of $\sim$40\% for both semiconductors at our 60$^{\circ}$ incidence angle, the photo-injected electron density is $\sim10^{18}cm^{-3}$.
At these carrier densities, which are non-degenerate for the upper $\Gamma_8$ conduction bands, we expect rapid thermalization \cite{portella_k-space_1992} and, consequently, an electron distribution with a well-populated Boltzmann tail extending above the vacuum level located at an effective work function $\Phi$ of 0.99eV (1.18eV) above the $\Gamma_8$ conduction band minimum in GaSb (InSb); thus allowing for thermionic emission from this excited state as initially $\exp[\Phi/(k_B T_e)] \approx 0.06$ in both semiconductors.

Carrier cooling, primarily by longitudinal optical (LO) phonon emission, and decay out of the upper conduction band to lower states will rapidly deplete the Boltzmann tail above the effective $\sim$1eV work function, thereby curtailing the observed ESTE.
By comparison with the $\Gamma_6$ conduction band Fr\"ohlich coupling constant in GaAs, where the characteristic LO phonon emission time is 165fs for a hot electron in the $\Gamma_6$ condunction band \cite{kash_subpicosecond_1985}, we estimate that hot electrons in the upper $\Gamma_8$ conduction band of both GaSb and InSb emit a LO phonon every $\sim$200fs with an energy of 29 and 24meV, respectively.
This means that even without population decay mechanisms the temperature of the exited electron distribution $T_e$ drops at a rate of $\sim$1,600K/ps in both semiconductors.
We note that the close proximity of the lower and unconfined (in momentum space) $\Gamma_7$ upper conduction band to the $\Gamma_8$ band will likely result in a fast population decay, so that the observed ESTE process should have an intrinsic latency of much less than the $\sim$4ps UV laser pulse duration.

\ref{fig:este-semicond} also shows that the measured Fourier plane electron beam spot size for GaSb is independent of the UV laser pulse energy, as would be expected for the proposed emission process.
More significantly, using $\Delta p_{\smallT} = \sqrt{m k_B T_e}$ (i.e., assuming the emitted electrons have a mass $m$ in the semiconductor), simulation of the experiment with the extended AG model \cite{berger_semi-analytic_2010} indicates that the observed GaSb Fourier plane spot size of 0.90($\pm$0.06)mm (HWe$^{-1}$M) would be due to an electron temperature $T_e \approx$ 360K which is associated with negligible thermionic emission ($\exp[-\Phi/(k_B T_e)] \sim 10^{-15}$).
On the other hand, employing a mass of 0.3$m$ for the electrons in GaSb allows an average electron temperature (over the $\sim$4ps laser pulse duration) of about 1,200K to be extracted using the AG model simulation --- a value much more consistent with the proposed ESTE mechanism and the expected cooling rate by LO phonon emission.
In fact, at the maximum $\sim$0.2nJ incident UV laser pulse energy up to $10^8$ electrons/pulse could be excited into the upper $\Gamma_8$ conduction band, of which one in $\exp[\Phi/(k_B T_e)] \approx$ 20,000 are above the effective 0.99eV work function, thus agreeing with the observed yield of $\sim$10$^3$ electrons/pulse (\ref{fig:este-semicond}).
We also note that the normalized transverse rms emittance $\varepsilon_{\smallT}$ of this UV laser-driven ultrafast GaSb electron source is more than a factor of two less than that expected from a standard Cu photocathode ($\Phi$ = 4.31eV and assuming $\Delta p_{\smallT} = \sqrt{m ( \hbar \omega - \Phi ) / 3 }$, where $\hbar \omega$ is the incident photon energy \cite{dowell_quantum_2009,jensen_emittance_2010}) irradiated at the same 261nm wavelength.

Very similar results are observed with the $p$-type InSb photocathode except that the Fourier plane spot size is 30-40\%  larger and the electron yield is a factor of 1.7 higher.
The increased rms transverse momentum for the same ESTE process is consistent with the larger $\sim$0.5$m$ effective mass of the upper $\Gamma_8$ conduction band in InSb, the expected higher initial temperature of the electron distribution, and the marginally slower cooling process associated with the lower 24meV LO phonon energy.
The resultant higher average electron temperature $T_e$ also contributes to the increased electron yield that is likely further enhanced by a larger absorption into the upper $\Gamma_8$ conduction band (due to its larger effective mass) and a 4\% lower $p$-polarized reflectance for the 60$^{\circ}$ incident UV laser radiation \cite{aspnes_dielectric_1983}.

The dependence of $\Delta p_{\smallT}$ on $m^*$ is readily explained through a consideration of energy and momentum conservation in transmission across a boundary with a potential step associated with the effective photoelectric work function $\Phi$.
Inside the material, an electron with an energy $E$ above the Fermi energy which is greater than $\Phi$ has a maximum momentum of $\sqrt{2 m^* (E-\Phi) }$ parallel to the boundary if it is to be emitted.
As this momentum component is conserved in emission from the material surface into the vacuum, the rms transverse momentum $\Delta p_{\smallT}$ of the electron source clearly must scale with $\sqrt{m^*}$ as observed in the experiment.
This is, of course, the \textit{sine qua non} of angle-resolved photoemission studies \cite{himpsel_angle-resolved_1983} where determination of the electron emission angle and energy allows the effective mass parallel to the sample surface to be determined.
We note that the electron effective mass also causes the narrow emission cone reported for $p$-type GaAs(100) negative electron affinity photocathodes \cite{liu_narrow_2005}, where surface cesiation lowers the vacuum level below the $\Gamma_6$-valley minimum ($m^* = 0.067m$)  to allow for direct emission of electrons excited into the conduction band.

>>>>

Recent experiments in our group have come to change our understanding of several photoemission processes, once thought to be well understood~\cite{berger_excited_2012}.
Indeed, while this knowledge has been commonplace in other fields, such as angle-resolved photoemission spectroscopy (ARPES)~\cite{himpsel_angle-resolved_1983}, to our knowledge the effect has not been considered for electron beam generation.
In this section, we show that, at least for thermionic emission, the effective mass $m^*$ of the electronic state from which the electron was emitted affects the rms transverse momentum; that is, the expression for $\Delta p_{\smallT}$ developed in Refs.~\cite{dowell_quantum_2009} and~\cite{jensen_emittance_2010} for thermionic electron emission should be rewritten to read $\sqrt{m^* k_B T_e}$, where $T_e$ is the temperature of the emitting electron distribution and $k_B$ is Boltzmann's constant.
Direct observation of the transverse momentum distribution for excited-state thermionic emission (ESTE) from GaSb and InSb, two similar III-V zincblende semiconductors, demonstrates the $m^*$ dependence.
Simulations of the experiment using our extended analytical Gaussian (AG) model of electron pulse propagation (see Section \ref{sec:external_forces}) are in close agreement with the data.

@@

\ref{fig:este-semicond} depicts the results obtained for GaSb and InSb photocathode materials under pulsed UV (261nm) laser irradiation using the same experimental technique employed for PAPE, 2PTE, and single-photon PE studies.
Both samples are cut from [100]-oriented polished wafers; the GaSb is undoped and the InSb is moderately $p$-type.
Prior studies~\cite{gobeli_photoelectric_1965} indicate that single-photon photoemission should not possible for either semiconductor --- the effective photoemission work functions (about 4.8eV for undoped GaSb and 4.89eV for $p$-type InSb (Fermi level pinned at valence band maximum)) being greater than the 4.75eV photon energy even when the 34meV Schottky barrier suppression due to the applied 8kV/cm DC field is included~\cite{dowell_quantum_2009}.
Nonetheless, for both zincblende semiconductors, significant laser-driven emission (more than expected for a single-photon field emission process) is observed with a yield nearly linearly proportional to the $\sim$4ps UV laser pulse energy.
Note that due to the surprisingly large yield, we repeated the measurements, this time replacing the the scintillator screen with a Faraday cup and repeating the laser energy sequence.
In this way, the yields for these photocathodes were measured directly and absolutely, unlike the results presented in the preceeding sections, for which relative measurements were sufficient.

The known optical properties of the two semiconductors~\cite{aspnes_dielectric_1983} indicate a strong absorption at 261nm, with an optical absorption depth of only 7-8nm, which, based on the band structures of GaSb~\cite{chelikowsky_nonlocal_1976} and InSb~\cite{chelikowsky_erratum_1984}, is primarily due to promotion of electrons near the $\Gamma$ point from the valence band (heavy-hole, light-hole, and split-off bands) directly into the upper $\Gamma_8$ conduction band located at 3.77eV and 3.59eV above the valence band maximum in GaSb and InSb, respectively.
Assuming parabolic bands and an estimated effective mass $m^*$ of about 0.3$m$ (0.5$m$) for the $\Gamma_8$ conduction band in GaSb (InSb), we determine that these electrons are excited with an average excess energy (above the $\Gamma_8$ conduction band minimum) of $\sim$0.35eV ($\sim$0.41eV); that is, an initial electron temperature $T_e$ of $\sim$4,200K ($\sim$4,900K).
For the $\sim$0.1nJ incident UV pulse energies, $\sim40\mu$m laser spot size, and $p$-polarized reflectance of $\sim$40\% for both semiconductors at our 60$^{\circ}$ incidence angle, the photo-injected electron density is $\sim10^{18}cm^{-3}$.
At these carrier densities, which are non-degenerate for the upper $\Gamma_8$ conduction bands, we expect rapid thermalization~\cite{portella_k-space_1992} and, consequently, an electron distribution with a well-populated Boltzmann tail extending above the vacuum level located at an effective work function $\Phi$ of 0.99eV (1.18eV) above the $\Gamma_8$ conduction band minimum in GaSb (InSb); thus allowing for thermionic emission from this excited state as initially $\exp[\Phi/(k_B T_e)] \approx 0.06$ in both semiconductors.

Carrier cooling, primarily by longitudinal optical (LO) phonon emission, and decay out of the upper conduction band to lower states will rapidly deplete the Boltzmann tail above the effective $\sim$1eV work function, thereby curtailing the observed ESTE.
By comparison with the $\Gamma_6$ conduction band Fr\"ohlich coupling constant in GaAs, where the characteristic LO phonon emission time is 165fs for a hot electron in the $\Gamma_6$ condunction band~\cite{kash_subpicosecond_1985}, we estimate that hot electrons in the upper $\Gamma_8$ conduction band of both GaSb and InSb emit a LO phonon every $\sim$200fs with an energy of 29 and 24meV, respectively.
This means that even without population decay mechanisms the temperature of the exited electron distribution $T_e$ drops at a rate of $\sim$1,600K/ps in both semiconductors.
We note that the close proximity of the lower and unconfined (in momentum space) $\Gamma_7$ upper conduction band to the $\Gamma_8$ band will likely result in a fast population decay, so that the observed ESTE process should have an intrinsic latency of much less than the $\sim$4ps UV laser pulse duration.

\ref{fig:este-semicond} also shows that the measured Fourier plane electron beam spot size for GaSb is independent of the UV laser pulse energy, as would be expected for the proposed emission process.
More significantly, using $\Delta p_{\smallT} = \sqrt{m k_B T_e}$ (i.e., assuming the emitted electrons have a mass $m$ in the semiconductor), simulation of the experiment with the extended AG model~\cite{berger_semi-analytic_2010} indicates that the observed GaSb Fourier plane spot size of 0.90($\pm$0.06)mm (HWe$^{-1}$M) would be due to an electron temperature $T_e \approx$ 360K which is associated with negligible thermionic emission ($\exp[-\Phi/(k_B T_e)] \sim 10^{-15}$).
On the other hand, employing a mass of 0.3$m$ for the electrons in GaSb allows an average electron temperature (over the $\sim$4ps laser pulse duration) of about 1,200K to be extracted using the AG model simulation --- a value much more consistent with the proposed ESTE mechanism and the expected cooling rate by LO phonon emission.
In fact, at the maximum $\sim$0.2nJ incident UV laser pulse energy up to $10^8$ electrons/pulse could be excited into the upper $\Gamma_8$ conduction band, of which one in $\exp[\Phi/(k_B T_e)] \approx$ 20,000 are above the effective 0.99eV work function, thus agreeing with the observed yield of $\sim$10$^3$ electrons/pulse (\ref{fig:este-semicond}).
We also note that the normalized transverse rms emittance $\varepsilon_{\smallT}$ of this UV laser-driven ultrafast GaSb electron source is more than a factor of two less than that expected from a standard Cu photocathode ($\Phi$ = 4.31eV and assuming $\Delta p_{\smallT} = \sqrt{m ( \hbar \omega - \Phi ) / 3 }$, where $\hbar \omega$ is the incident photon energy~\cite{dowell_quantum_2009,jensen_emittance_2010}) irradiated at the same 261nm wavelength.

Very similar results are observed with the $p$-type InSb photocathode except that the Fourier plane spot size is 30-40\%  larger and the electron yield is a factor of 1.7 higher.
The increased rms transverse momentum for the same ESTE process is consistent with the larger $\sim$0.5$m$ effective mass of the upper $\Gamma_8$ conduction band in InSb, the expected higher initial temperature of the electron distribution, and the marginally slower cooling process associated with the lower 24meV LO phonon energy.
The resultant higher average electron temperature $T_e$ also contributes to the increased electron yield that is likely further enhanced by a larger absorption into the upper $\Gamma_8$ conduction band (due to its larger effective mass) and a 4\% lower $p$-polarized reflectance for the 60$^{\circ}$ incident UV laser radiation~\cite{aspnes_dielectric_1983}.

The dependence of $\Delta p_{\smallT}$ on $m^*$ is readily explained through a consideration of energy and momentum conservation in transmission across a boundary with a potential step associated with the effective photoelectric work function $\Phi$.
Inside the material, an electron with an energy $E$ above the Fermi energy which is greater than $\Phi$ has a maximum momentum of $\sqrt{2 m^* (E-\Phi) }$ parallel to the boundary if it is to be emitted.
As this momentum component is conserved in emission from the material surface into the vacuum, the rms transverse momentum $\Delta p_{\smallT}$ of the electron source clearly must scale with $\sqrt{m^*}$ as observed in the experiment.
This is, of course, the \textit{sine qua non} of angle-resolved photoemission studies~\cite{himpsel_angle-resolved_1983} where determination of the electron emission angle and energy allows the effective mass parallel to the sample surface to be determined.
We note that the electron effective mass also causes the narrow emission cone reported for $p$-type GaAs(100) negative electron affinity photocathodes~\cite{liu_narrow_2005}, where surface cesiation lowers the vacuum level below the $\Gamma_6$-valley minimum ($m^* = 0.067m$)  to allow for direct emission of electrons excited into the conduction band.

### 2013-04-11 inc/photocathode/future.tex
citations are now preceeded by non-breaking space (~)

The philosophical argument presented in Section \ref{sec:este} for using the effective mass $m^*$ in the expressions for the transverse momentum variance $\Delta p_{\smallT}$ simply follows from conservation of energy and momentum at the barrier and actually does not depend on the ``excited state'' nature nor the thermionic emission process.
For the example of two-photon assisted thermionic emission (2PTE) from gold presented in Section \ref{sec:two_photon_thermionic}, the close agreement between the data and the simulation is supported by the fact that the effective mass of a free electron in Au is equal to its rest mass $m$ \cite{johnson_optical_1972} --- the value employed in that analysis.

For direct single-photon photoemission, we would therefore also expect to require that the expression describing the rms transverse momentum (\ref{eq:jensen_eta}, Refs. \cite{dowell_quantum_2009} and \cite{jensen_emittance_2010}) to be written as $\Delta p_{\smallT} = \sqrt{m^* ( \hbar \omega - \Phi ) / 3 }$.
Efforts are currently underway to determine if this is indeed the case using a variety of metal photocathodes and the $\sim$4ps, 261nm laser radiation source.
In an upcoming publication, we will show that the dependence of $\Delta p_{\smallT}$ on the electron effective mass may be extended to simple planar metal photocathodes --- thus potentially allowing for the future development or discovery of a robust, high brightness photocathode material.
For a selection of single-crystal and polycrystalline photocathodes, we will show that the simulation, incorporating $m^*$, is in good agreement with our direct measurements of $\Delta p_{\smallT}$.
We will also discuss the effect of laser-induced heating on $\Delta p_{\smallT}$ of the electron distribution in the photocathode during photoemission, which may limit the attainable pulsed electron source brightness at the desired high peak pulse current densities.

>>>>

The philosophical argument presented in Section \ref{sec:este} for using the effective mass $m^*$ in the expressions for the transverse momentum variance $\Delta p_{\smallT}$ simply follows from conservation of energy and momentum at the barrier and actually does not depend on the ``excited state'' nature nor the thermionic emission process.
For the example of two-photon assisted thermionic emission (2PTE) from gold presented in Section \ref{sec:two_photon_thermionic}, the close agreement between the data and the simulation is supported by the fact that the effective mass of a free electron in Au is equal to its rest mass $m$~\cite{johnson_optical_1972} --- the value employed in that analysis.

For direct single-photon photoemission, we would therefore also expect to require that the expression describing the rms transverse momentum (\ref{eq:jensen_eta}, Refs.~\cite{dowell_quantum_2009} and~\cite{jensen_emittance_2010}) to be written as $\Delta p_{\smallT} = \sqrt{m^* ( \hbar \omega - \Phi ) / 3 }$.
Efforts are currently underway to determine if this is indeed the case using a variety of metal photocathodes and the $\sim$4ps, 261nm laser radiation source.
In an upcoming publication, we will show that the dependence of $\Delta p_{\smallT}$ on the electron effective mass may be extended to simple planar metal photocathodes --- thus potentially allowing for the future development or discovery of a robust, high brightness photocathode material.
For a selection of single-crystal and polycrystalline photocathodes, we will show that the simulation, incorporating $m^*$, is in good agreement with our direct measurements of $\Delta p_{\smallT}$.
We will also discuss the effect of laser-induced heating on $\Delta p_{\smallT}$ of the electron distribution in the photocathode during photoemission, which may limit the attainable pulsed electron source brightness at the desired high peak pulse current densities.

### 2013-04-11 inc/photocathode/plasmon/plasmon.tex
citations are now preceeded by non-breaking space (~)

In the search for alternative mechanisms to reduce the intrinsic divergence of the electron beam upon photoemission, a possible area of interest was Plasmon-Assisted Photoemission (PAPE). 
As the name suggests, a plasmon is the quasiparticle of plasma oscillation.
When this occurs at the surface of a conducting material, the resultant quasiparticles are called ``surface plasmon-polaritons.''
Since these will be the only plasmons discussed in this thesis, the word ``plasmon'' will be used interchangeably for the name ``surface plasmon-polariton.''
Electrons in a conductor can couple with electromagnetic fields, in this case a laser field, and oscillate accordingly.
This overall oscillation of electrons near the surface of the conductor creates strong oscillatory local surface electric field \cite{cottam_introduction_2004,concepts_2002}.
In PAPE, the goal is use this field to both enhance the emission and `control' the divergence of photoemitted electrons through the combined action of photoemission barrier suppression and ponderomotive acceleration.
On the other hand, these oscillating electrons are nothing but an oscillating current, and thus if a large plasmon response is established, resistive heating is a major concern even for the small resistivities typical of conductors.

This avenue of investigation was motivated by Zawadzka et al \cite{zawadzka_evanescent_2001} who indicate that when driving a surface plasmon on gold they witnessed a reduction in the emission angle of photoemitted electrons.
In their paper, the authors coat a thin gold film on a prism; the laser then ``back illuminates'' the gold through the prism, ejecting electrons on vacuum side of the foil.
This arrangement is called the Kretschmann geometry.
While this geometry is common, due to fears about the lack of heat conduction through the glass and the power needed to generate a sufficient electron beam current, we felt that for UEM applications we should explore the alternative ``grating coupling'' geometry \cite{kupersztych_ponderomotive_2001,kupersztych_anomalous_2005,li_surface_2013}.
In this geometry, the plasmon is driven on a periodic surface rather than a flat surface attached to glass.
Futher, in this front-side illuminated geometry the back surface is now free to be attached to a heat sink.

@@

At the interface of a vacuum and such a simple material, contstraints on the material parameters which may allow surface plasmons, result in the condition that the dielectric constant at the oscillation frequency of the surface plasmon $ \varepsilon(\omega_{SP}) = -1 $ \cite{cottam_introduction_2004}.
Substituting this result into \ref{eq:dielectric_as_omega} shows that this frequency of oscillation is simply related to intrinsic plasma frequency by
\begin{equation}
  \omega_{SP} = \omega_{P} / \sqrt{2} \,\text{.}
\end{equation}
Further, plasmons propagating along a vacuum interface carry momentum whose magnitude is given by 
\begin{equation}
  k_{SP}^2 = \frac{\omega^2}{c^2} \left( \frac{ \varepsilon }{ 1 + \varepsilon } \right) \,\text{.}
\end{equation}

@@

Collaborators at Argonne National Laboratory's Electron Microscopy Center, Material Science Division employed a defocused Gallium focused ion beam (FIB) (Zeiss 1540XB) to mill an approximately sinusoidal shape into the surface of a silicon wafer.
Several regions were created each 100x100$\mu$m in area, having a $\sim$1$\mu$m period and $\sim$10\% modulation depth.
By \ref{eq:plasmon_condition}, given a laser wavelength of 523nm ($\varepsilon_{Au} \approx -3.95$ \cite{johnson_optical_1972}), this choice of periodicity corresponds to an incidence angle of $\sim39^{\circ}$, which is within the acceptance range of our experimental setup (35-75$^{\circ}$) (see Section \ref{sec:gun_design}).
An SEM image of one such region is shown in \ref{fig:fib-si-sem}.
A 300nm film of gold was deposited onto the silicon using the Varian e-beam deposition chamber at the UIC Nano Core Facility (NCF); a thin layer of chromium was deposited between the silicon and gold as a binding agent.
As the gold film is many times thicker than the penetration depth of green light on gold and several times thicker than the penetration depth of the plasmon field, neither the laser radiation nor the plasmon field will couple with the silicon.
%TODO should I comment on the coating acting to smooth the surface to a proper sinusoid?

@@

% Content taken from 2010 ANL report
\ref{fig:pape} summarizes the photoemission results obtained from this gold-coated periodic photocathode under green (523nm) laser irradiation.
Strong enhanced phototemission was observed for only a narrow $\pm$5mrad range of incidence angles $\theta$ around the expected plasmon resonance angle of 39$^{\circ}$.
Moreover, this enhanced laser-driven electron emission was only observed for $p$-polarized incident 523nm femtosecond laser radiation.
These two properties are known characteristics of plasmon-assisted photoemission \cite{kupersztych_ponderomotive_2001,kupersztych_anomalous_2005,li_surface_2013}.

@@

The experimental results are shown to the right in \ref{fig:rotated}.
As the divergence is clearly shown to be rotated by $45^{\circ}$ with respect to the colinear case, this experiment demonstrates that the cause of the divergence increase is due to the local field enhancements near the peaks of the grating.
For a mathematical treatment of the transverse electric field of such patterned surfaces, the reader is referred to Ref. \cite{watts_sharp_1997}. 
Our results are commensurate with the results shown by Zawadzka et al \cite{zawadzka_evanescent_2001}, who showed a reduced divergence in the Kretschmann geometry.
As that geometry does not employ a patterned surface, it would not be subject to such local field enhancements of the accelerating electric field.

@@

Other nano-engineered surfaces are still being considered for future work.
Various other more exotic nano-structures are under examination for PAPE \cite{li_surface_2013,polyakov_plasmon_2013}.
Bi-metallic layers are known to enhance PAPE, which may allow for reduced incident laser power for driving the plasmon \cite{kupersztych_anomalous_2005}.
Photoemission from nano-arrays of quantum-wells also offers the tantalizing hope of reduced rms transverse momentum of the generated electrons.

>>>>

In the search for alternative mechanisms to reduce the intrinsic divergence of the electron beam upon photoemission, a possible area of interest was Plasmon-Assisted Photoemission (PAPE). 
As the name suggests, a plasmon is the quasiparticle of plasma oscillation.
When this occurs at the surface of a conducting material, the resultant quasiparticles are called ``surface plasmon-polaritons.''
Since these will be the only plasmons discussed in this thesis, the word ``plasmon'' will be used interchangeably for the name ``surface plasmon-polariton.''
Electrons in a conductor can couple with electromagnetic fields, in this case a laser field, and oscillate accordingly.
This overall oscillation of electrons near the surface of the conductor creates strong oscillatory local surface electric field~\cite{cottam_introduction_2004,concepts_2002}.
In PAPE, the goal is use this field to both enhance the emission and `control' the divergence of photoemitted electrons through the combined action of photoemission barrier suppression and ponderomotive acceleration.
On the other hand, these oscillating electrons are nothing but an oscillating current, and thus if a large plasmon response is established, resistive heating is a major concern even for the small resistivities typical of conductors.

This avenue of investigation was motivated by Zawadzka et al~\cite{zawadzka_evanescent_2001} who indicate that when driving a surface plasmon on gold they witnessed a reduction in the emission angle of photoemitted electrons.
In their paper, the authors coat a thin gold film on a prism; the laser then ``back illuminates'' the gold through the prism, ejecting electrons on vacuum side of the foil.
This arrangement is called the Kretschmann geometry.
While this geometry is common, due to fears about the lack of heat conduction through the glass and the power needed to generate a sufficient electron beam current, we felt that for UEM applications we should explore the alternative ``grating coupling'' geometry~\cite{kupersztych_ponderomotive_2001,kupersztych_anomalous_2005,li_surface_2013}.
In this geometry, the plasmon is driven on a periodic surface rather than a flat surface attached to glass.
Futher, in this front-side illuminated geometry the back surface is now free to be attached to a heat sink.

@@

At the interface of a vacuum and such a simple material, contstraints on the material parameters which may allow surface plasmons, result in the condition that the dielectric constant at the oscillation frequency of the surface plasmon $ \varepsilon(\omega_{SP}) = -1 $~\cite{cottam_introduction_2004}.
Substituting this result into \ref{eq:dielectric_as_omega} shows that this frequency of oscillation is simply related to intrinsic plasma frequency by
\begin{equation}
  \omega_{SP} = \omega_{P} / \sqrt{2} \,\text{.}
\end{equation}
Further, plasmons propagating along a vacuum interface carry momentum whose magnitude is given by 
\begin{equation}
  k_{SP}^2 = \frac{\omega^2}{c^2} \left( \frac{ \varepsilon }{ 1 + \varepsilon } \right) \,\text{.}
\end{equation}

@@

Collaborators at Argonne National Laboratory's Electron Microscopy Center, Material Science Division employed a defocused Gallium focused ion beam (FIB) (Zeiss 1540XB) to mill an approximately sinusoidal shape into the surface of a silicon wafer.
Several regions were created each 100x100$\mu$m in area, having a $\sim$1$\mu$m period and $\sim$10\% modulation depth.
By \ref{eq:plasmon_condition}, given a laser wavelength of 523nm ($\varepsilon_{Au} \approx -3.95$~\cite{johnson_optical_1972}), this choice of periodicity corresponds to an incidence angle of $\sim39^{\circ}$, which is within the acceptance range of our experimental setup (35-75$^{\circ}$) (see Section \ref{sec:gun_design}).
An SEM image of one such region is shown in \ref{fig:fib-si-sem}.
A 300nm film of gold was deposited onto the silicon using the Varian e-beam deposition chamber at the UIC Nano Core Facility (NCF); a thin layer of chromium was deposited between the silicon and gold as a binding agent.
As the gold film is many times thicker than the penetration depth of green light on gold and several times thicker than the penetration depth of the plasmon field, neither the laser radiation nor the plasmon field will couple with the silicon.
%TODO should I comment on the coating acting to smooth the surface to a proper sinusoid?

@@

% Content taken from 2010 ANL report
\ref{fig:pape} summarizes the photoemission results obtained from this gold-coated periodic photocathode under green (523nm) laser irradiation.
Strong enhanced phototemission was observed for only a narrow $\pm$5mrad range of incidence angles $\theta$ around the expected plasmon resonance angle of 39$^{\circ}$.
Moreover, this enhanced laser-driven electron emission was only observed for $p$-polarized incident 523nm femtosecond laser radiation.
These two properties are known characteristics of plasmon-assisted photoemission~\cite{kupersztych_ponderomotive_2001,kupersztych_anomalous_2005,li_surface_2013}.

@@

The experimental results are shown to the right in \ref{fig:rotated}.
As the divergence is clearly shown to be rotated by $45^{\circ}$ with respect to the colinear case, this experiment demonstrates that the cause of the divergence increase is due to the local field enhancements near the peaks of the grating.
For a mathematical treatment of the transverse electric field of such patterned surfaces, the reader is referred to Ref.~\cite{watts_sharp_1997}. 
Our results are commensurate with the results shown by Zawadzka et al~\cite{zawadzka_evanescent_2001}, who showed a reduced divergence in the Kretschmann geometry.
As that geometry does not employ a patterned surface, it would not be subject to such local field enhancements of the accelerating electric field.

@@

Other nano-engineered surfaces are still being considered for future work.
Various other more exotic nano-structures are under examination for PAPE~\cite{li_surface_2013,polyakov_plasmon_2013}.
Bi-metallic layers are known to enhance PAPE, which may allow for reduced incident laser power for driving the plasmon~\cite{kupersztych_anomalous_2005}.
Photoemission from nano-arrays of quantum-wells also offers the tantalizing hope of reduced rms transverse momentum of the generated electrons.
